---
title: "p8105_hw2_jy3360"
author: "Candice Yu"
date: "2023-09-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = F}
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```

# Problem 1
First, clean the data in pols-month.csv. 

- Use separate() to break up the variable mon into integer variables year, month, and day; 
- replace month number with month name; 
- create a president variable taking values gop and dem, 
- and remove prez_dem and prez_gop; and remove the day variable.

```{r}
library(tidyverse)
pols_month <- read_csv("fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year), day = as.integer(day)) %>%
  mutate(month = recode(month, `01` = "Jan", `02` = "Feb", `03` = "Mar", 
                        `04` = "Apr", `05` = "May", `06` = "June", 
                        `07` = "July", `08` = "Aug", `09` = "Sep", 
                        `10` = "Oct", `11` = "Nov", `12` = "Dec")) %>%
  mutate(president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) %>%
  select(-c("day", "prez_gop", "prez_dem"))
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, 
- arrange according to year and month, 
- and organize so that year and month are the leading columns.

```{r}
snp <- read_csv("fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  separate(date, into = c("year", "month", "day"), sep = "-") %>%
  mutate(year = as.integer(year), day = as.integer(day), month = as.integer(month)) %>%
  arrange(year, month) %>%
  relocate(year, month) %>%
  mutate(month = recode(month, `01` = "Jan", `02` = "Feb", `03` = "Mar", 
                        `04` = "Apr", `05` = "May", `06` = "June", 
                        `07` = "July", `08` = "Aug", `09` = "Sep", 
                        `10` = "Oct", `11` = "Nov", `12` = "Dec")) %>%
  select(-"day")

```

Third, tidy the unemployment data so that it can be merged with the previous datasets.

- This process will involve switching from “wide” to “long” format; 
- ensuring that key variables have the same name; 
- and ensuring that key variables take the same values.
```{r}
unemployment <- read_csv("fivethirtyeight_datasets/unemployment.csv") %>%
  pivot_longer(Jan:Dec, 
               names_to = "month",
               values_to = "unemployment"
               ) %>%
  janitor::clean_names()
```

Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r}
data_1 <- left_join(pols_month, snp, by = c("year", "month")) %>%
  left_join(unemployment, by = c("year", "month"))
str(data_1)
```


Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.

`r bfcolor("Answer 1:", "blue")` 

- There are `r nrow(pols_month)` rows and `r ncol(pols_month)` columns in the `pols_month` dataset. 
- There are `r nrow(snp)` rows and `r ncol(snp)` columns in the `snp` dataset. 
- There are `r nrow(unemployment)` rows and `r ncol(unemployment)` columns in the `unemployment` dataset. 
- There are `r nrow(data_1)` rows and `r ncol(data_1)` columns in the merged dataset `data_1`. 
- The years range from `r range(pols_month$year)[1]` to `r range(pols_month$year)[2]` in the merged dataset `pols_month`. 
- The years range from `r range(snp$year)[1]` to `r range(snp$year)[2]` in the merged dataset `snp`.
- The years range from `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]` in the merged dataset `unemployment`.
- The years range from `r range(data_1$year)[1]` to `r range(data_1$year)[2]` in the merged dataset `data_1`. 
- `pols_month`: It contains the data on the political party association of the president. The key variables contain `r names(pols_month)`.
- `snp`: It contains the data related to the S&P stock market index. The key variables contain `r names(snp)`.
- `unemployment`: It contains the data on the monthly unemployment rate. The key variables contain `r names(unemployment)`.
- `data_1`: The key variables contain `r names(data_1)`.

# Problem 2
Read and clean the Mr. Trash Wheel sheet:

- specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
- use reasonable variable names
- omit rows that do not include dumpster-specific data
- The data include a column for the (approximate) number of homes powered. This calculation is described in the Homes powered note, but not applied to every row in the dataset. Update the data to include a new homes_powered variable based on this calculation.
```{r}
library(readxl)
mr_trash_wheel <- readxl::read_excel("202309 Trash Wheel Collection Data.xlsx", 
                                  sheet = "Mr. Trash Wheel", 
                                  range = "A2:N586") %>%
  janitor::clean_names() %>%
  mutate(year = as.integer(year)) %>%
  mutate(homes_powered = weight_tons*500/30) %>%
  mutate(trash_wheel = "mr") %>%
  drop_na(dumpster) 
```

Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda,

- and combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. 
- To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.
```{r}
professor_trash_wheel <- readxl::read_excel("202309 Trash Wheel Collection Data.xlsx", 
                                  sheet = "Professor Trash Wheel", 
                                  range = "A2:M108") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(homes_powered = weight_tons*500/30) %>%
  mutate(trash_wheel = "professor")

gwynnda_trash_wheel <- readxl::read_excel("202309 Trash Wheel Collection Data.xlsx", 
                                  sheet = "Gwynnda Trash Wheel", 
                                  range = "A2:L157") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(homes_powered = weight_tons*500/30) %>%
  mutate(trash_wheel = "gwynnda")

trash_wheel_total <- bind_rows(mr_trash_wheel, professor_trash_wheel, gwynnda_trash_wheel) %>%
  relocate(trash_wheel)
```

Write a paragraph about these data; you are encouraged to use inline R.

- Be sure to note the number of observations in the resulting dataset, and give examples of key variables. 
- For available data, what was the total weight of trash collected by Professor Trash Wheel? 
- What was the total number of cigarette butts collected by Gwynnda in July of 2021?

`r bfcolor("Answer 2:", "blue")` 

- `mr_trash_wheel` contains `r nrow(mr_trash_wheel)` rows and `r ncol(mr_trash_wheel)` columns of data. 
- `professor_trash_wheel` contains `r nrow(professor_trash_wheel)` rows and `r ncol(professor_trash_wheel)` columns of data. 
- `gwynnda_trash_wheel` contains `r nrow(gwynnda_trash_wheel)` rows and `r ncol(gwynnda_trash_wheel)` columns of data. 
- The final data set `trash_wheel_total` contains `r nrow(trash_wheel_total)` rows and `r ncol(trash_wheel_total)` columns of data. The variables are `r names(trash_wheel_total)`.
- The total weight of trash collected by Professor Trash Wheel is `r sum(professor_trash_wheel$weight_tons)`. 
- The total number of cigarette butts collected by Gwynnda in July of 2021 is `r as.integer(sum(gwynnda_trash_wheel$cigarette_butts[which(gwynnda_trash_wheel$month == "July" & gwynnda_trash_wheel$year == 2021)]))`. 

# Problem 3

Import, clean, and tidy the dataset of baseline demographics. 

- Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), 
- and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 
```{r}
baseline_original <- read_csv("data_mci/MCI_baseline.csv", skip = 1, na = ".") %>%
  janitor::clean_names() %>%
  mutate(sex = recode(sex, `0` = "female", `1` = "male")) %>%
  mutate(apoe4 = recode(apoe4, `0` = "non-carrier", `1` = "carrier")) 

baseline <- read_csv("data_mci/MCI_baseline.csv", skip = 1) %>%
  janitor::clean_names() %>%
  mutate(sex = recode(sex, `0` = "female", `1` = "male")) %>%
  mutate(apoe4 = recode(apoe4, `0` = "non-carrier", `1` = "carrier")) %>%
  filter(age_at_onset!= ".")

```
- Discuss important steps in the import process and relevant features of the dataset. 
- How many participants were recruited, and of these how many develop MCI? 
- What is the average baseline age? 
- What proportion of women in the study are APOE4 carriers?


`r bfcolor("Answer 2:", "blue")` 

- During the import process, it's important to skip the first line of the data since it is the dictionary of the variable names. 
- The variables are `r names(baseline)`.
- `r nrow(read_csv("data_mci/MCI_baseline.csv", skip = 1))` participants were recruited. Of those `r nrow(read_csv("data_mci/MCI_baseline.csv", skip = 1))` participants, `r nrow(baseline)` developed MCI. 
- The average baseline age is `r mean(baseline$current_age)`. 
- The proportion of women in the study who are APOE4 carriers is `r (count(baseline_original[which(baseline_original$sex == "female" & baseline_original$apoe4 == "carrier"),]))/(count(baseline_original[which(baseline_original$sex == "female"),]))`


Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.
```{r}
library(tidyverse)
amyloid <- read_csv("data_mci/mci_amyloid.csv", skip = 1, na = c("Na", "NA")) %>%  # import the data, skip the first line, correct format of the missing values
  janitor::clean_names() %>%       # clean the variable names 
  pivot_longer(
    baseline:time_8,
    names_to = "time", # new single column for columns baseline to time_8
    names_prefix = "time_", # skip the prefix 
    values_to = "amyloid" # new single column for observed biomarker values
    ) %>%
  mutate(time = recode(time, "baseline" = "0")) %>%      # change the baseline to 0 in time
  mutate(time = as.integer(time))         # change all time to integers 
  
```


`r bfcolor("Answer 3.1:", "blue")` 

- The biomarker amyloid dataset includes information on the subject time(in years) elapsed since the study baseline to the visit where biomarker amyloid 42/40 ratio was measured. 
- `amyloid` includes `r length(unique(amyloid$time))` records at time `r as.integer(unique(amyloid$time))`. The dataset has `r nrow(amyloid)` rows and `r ncol(amyloid)` columns.
- `amyloid` includes variables: `r names(amyloid)`


Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. 
```{r}
# Participants appear only in the baseline dataset but not in the amyloid dataset
subset(baseline_original, !(id %in% amyloid$study_id))

# Participants appear only in the amyloid dataset but not in the baseline dataset 
subset(amyloid, !(study_id %in% baseline_original$id))
```

`r bfcolor("Answer 3.2:", "blue")` 

- There are `r length(subset(baseline_original, !(id %in% amyloid$study_id)))` of participants appear only in the baseline dataset but not in the amyloid dataset.
- There are `r length(unique((subset(amyloid, !(study_id %in% baseline_original$id)))$study_id))` of participants appear only in the amyloid dataset but not in the baseline dataset. 
- Participants appear only in the `amyloid` dataset but not in the baseline dataset has a study_id greater than or equal to 484. 



Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, 

- and briefly describe the resulting dataset; 
- export the result as a CSV to your data directory.

```{r}
amyloid_baseline <- inner_join(amyloid, baseline_original, by = join_by("study_id" == "id"))
```

`r bfcolor("Answer 3.3:", "blue")` 

- The combined dataset `amyloid_baseline` contains `r nrow(amyloid_baseline)` observations and `r ncol(amyloid_baseline)` variables. 
- There are `r length(unique(amyloid_baseline$study_id))` participants who appear in both datasets. 
- The dataset includes variables are `r names(amyloid_baseline)`.

```{r}
write_csv(amyloid_baseline, "data_mci/amyloid_baseline.csv")  # export the result as a CSV to your data directory
```





